XRONOS: an open data infrastructure for archaeological chronology

### 2024-11-24 paper_jcaa.qmd
updated Dendro Part and turned it into prose

<!-- TODO: MH, update and turn into prose -->
For European dendrochronological data, the situation is even less favourable for archaeological purposes:

* NOAA stores some global datasets, but has its main focus in the United States and, with its only 34 European prehistoric datasets, is not useful for the study of the European past.
* The ADS database is compiled and maintained by the UK-based Vernacular Architecture Group and is limited to data published annually in the journal Vernacular Architecture, containing only medieval or later dates for the UK.
* DendroDB has a European focus but only provides data for historical periods.
* The "Dendrochronological Database" of the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) has a worldwide focus on collected natural wood samples and has different goals than a database for archaeological wood material (furthermore, although announced a long time ago, this database is still not functional).
* Although "Digital Collaboratory for Cultural Dendrochronology (DCCD). An international digital data library for dendrochronology" is well connected in the context of archaeological data services (e.g. ARIADNE), it is focused on the Netherlands (more than 2/3 of the data, while 0.08% of the Quercus data refer to Switzerland) and contains very few prehistoric data (about 2.5%). Furthermore, the overall activity in this database has decreased from 3846 new project records in 2010-2014 to only 83 since then (until the end of 2019).

>>>>

<!-- TODO: MH, update and turn into prose. From MH: Should be done now, OK like this? -->
For European dendrochronological data, the situation is even less favourable for archaeological purposes:

Some databases provide valuable resources for dendrochronological studies in general but are not primarily designed to collect tree-ring data from archaeological contexts. For instance, Dendro4Art specializes in dendrochronological data related to wooden art objects, such as sculptures and panel paintings. While this focus serves art historians and conservationists well, its utility for studying prehistoric datasets is minimal. Similarly, the Dendrochronological Picture Database, maintained by the Swiss Federal Institute for Snow and Avalanche Research (SLF), offers a visual archive of approximately 1,400 images documenting dendrochronological phenomena. Although valuable as an educational resource, it does not provide raw data necessary for chronological or archaeological analyses. Additionally, the OLDLIST and Eastern OLDLIST databases focus on documenting the maximum ages of trees worldwide. Their emphasis on biological longevity, while significant for ecological research, limits their applicability to archaeological or prehistoric investigations.

Among databases that do provide dendrochronological data, the degree to which they support prehistoric research varies substantially. The NOAA International Tree-Ring Data Bank (ITRDB) serves as a global repository of tree-ring measurements. However, its focus remains predominantly on North America, with only 34 datasets representing European prehistoric contexts. This restricts its relevance for studying the European past. Similarly, the ADS database, maintained by the UK-based Vernacular Architecture Group, compiles dendrochronological data from the UK but is limited to medieval and later periods, making it unsuitable for prehistoric studies.

DendroDB, hosted by the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL), emphasizes ecological and climate studies over archaeological wood material. While it claims a broad scope, the database remains non-functional, rendering it ineffective for research needs. The CFS-TRenD database, managed by the Canadian Forest Service, compiles over 4,600 datasets from Canadian forests, primarily focusing on boreal ecosystems. Despite its extensive coverage for North America, its geographical specificity and lack of open access restrict its utility for European prehistoric contexts. Similarly, the QUB Dendrochronology Database, managed by Queen’s University Belfast, offers valuable datasets for Ireland and the UK but lacks significant representation of prehistoric material, limiting its application in broader archaeological investigations. The Building Archaeology Research Database (BARD) contains over 24,000 records, including dendrochronological data from more than 2,700 buildings. However, its focus on medieval and post-medieval timber-frame construction further narrows its utility for studies involving prehistoric wood samples.

The Digital Collaboratory for Cultural Dendrochronology (DCCD) presents itself as a potentially valuable international platform for dendrochronology, particularly through its integration with archaeological data services such as ARIADNE. However, it remains heavily biased toward datasets from the Netherlands, which account for more than two-thirds of its entries, while only 0.08% of its Quercus data pertain to Switzerland and just 2.5% represent prehistoric datasets. Notably, these estimates date back to 2021, and an updated assessment is currently unattainable following the platform’s migration to DataverseNL, which now charges an annual fee of nearly €6,000 for access. Furthermore, database activity has declined significantly, from 3,846 new project records between 2010 and 2014 to only 83 by the end of 2019. Although 519 additional records have been reported since 2021, it remains unclear whether this figure includes revisions to pre-existing entries, potentially inflating the count. The database’s narrow focus and restrictive access model significantly limit its broader utility for prehistoric research.

Finally, the Strategic Environmental Archaeology Database (SEAD), hosted by Umeå University, integrates multiple environmental proxies, including dendrochronological data. However, the dendrochronological component is largely confined to Swedish data, with limited relevance to prehistoric contexts. While SEAD aims for broader applications, its dendro component has limited utility for studies outside Sweden.

In conclusion, the utility of dendrochronological databases for prehistoric research varies widely. Global resources such as NOAA’s ITRDB and DCCD offer substantial datasets but face significant limitations in practical geographical and temporal scope. Similarly, platforms like DendroDB and BARD primarily cater to historical studies, leaving critical gaps in prehistoric coverage. Specialized resources like OLDLIST, Dendro4Art, and the Dendrochronological Picture Database provide valuable contributions but lack the direct relevance necessary for archaeological tree-ring analysis. Consequently, researchers focused on prehistoric dendrochronology must navigate a fragmented landscape of databases, each offering distinct strengths and limitations. Addressing these gaps remains crucial for advancing the field.
